---
title: "LPWC: Lag Penalized Weighted Correlation for Clustering Short Time Series"
author: "Thevaa Chandereng & Anthony Gitter"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::2'"
abstract: > 
  
 
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_float: true
    fig_width: 5
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Analyzing RNA-seq data with DESeq2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png",
                      message=FALSE, error=FALSE, warning=TRUE)
```	


# Introduction

Time series data unlike other data are a series of data collected across different timepoints which are dependent on the past data.
In biology, most time series data consist of fewer time points making the analysis of time series data more complicated.
Good clustering method in time series captures similar temporal shape in a dataset. Similarity in expression patterns can correspond to similarity in biological function, which helps aid in the direction of future research. 
Lag Penalized Weighted Correlation(LPWC) is a distance based clustering permiting lags using weighted correlation. 
LPWC accomodates lags but due to the length of the time series, the lags are penalized using gausssian kernel. 
Unlike many other time series data, LPWC also accounts for irregular time sampling.


If you use LPWC in published research, please cite:


If after reading through this vignette you have questions or problems
using LPWC, please post them to https://support.bioconductor.org and tag "LPWC".
This will notify the package maintainers and benefit other users.


You should **not** email your question to the package authors, as we will
just reply that the question should be posted to the 
**Bioconductor support site**.



# Run LPWC

Prior to analyzing your data, the R package needs to be installed

Since, the data is not published to Bioconductor/CRAN. The best way to download the R package through Github

```{r install, results='asis', eval=FALSE}
library(devtools)
devtools::install_github("gitter-lab/LPWC")
```

After successful installation, the package must be loaded into the working space:

```{r lib, results="asis", eval=TRUE}
library(LPWC)
```



## Required Input

**Data :**  Input to LPWC is a matrix. The expression matix shoud be n -by- p where n is the number of genes (in rows) and p is the timepoints where data where collected. 

The object **simdata** is a simulated data for 200 genes with 8 timepoints. This is stored as a matrix.



```{r data, results='markup'}
data(simdata)
simdata[1:5, ]
str(simdata)
```



**Timepoints: ** The object **timepoints** should be a vector of timepoints where data was collected. This should match column length in **Data**. This can be in any form (secs, mins, hours or even days), but make sure they are uniform throughout. 

The timepoints used in **Data** are
```{r time}
timepoints <- c(0, 2, 4, 6, 8, 18, 24, 32, 48, 72)
timepoints
```


**Data** and **Timepoints** are the main requirement to run LPWC, however, there are other parameters that is used in running LPWC. These are not required, but they do affect the results of LPWC

**Max.lag: ** The object **max.lag** should be a numerical output where the maximum lag is specified. This max.lag has to be less than the floor of length(timepoints) / 4. 


**C:** The object **C** should be a numeric. This controls the gaussian kernel function used to impose penalty on the lag and weights in weighed correlation vector. For more details, please look at the literaure. 


**penalty:** The object **penalty** only takes two arguments: low and high. High penalty imposes higher penalty on lags thus favoring non-lag cases, meanwhile lower peanlty does vice versa. The default is set to "high" penalty. For more details, please look at the literaure. 


**iter:** The object *iter* should be a numeric. This controls the number of C should be tested. The default is set at 10, but you can increase or decrease it. Increasing iter, increases computational time. This is only relevant for low penalty as the high penalty have a fixed penalty. 


## LPWC : Similarity

LPWC will compute similarity/correlation between genes and return three output: correlation of genes in dist form (see as.dist), lags used for each gene and the C used in the LPWC. 


An example output of LPWC for 10 genes using high penalty


```{r}

LPWC::corr.bestlag(simdata[49:58, ], timepoints = timepoints, max.lag = 2, penalty = "high", iter = 10)
```

However, most clustering algorithm takes distance as an ouput, especially hierarchical clustering, we can convert the similarity to distance by subtracting the similarity from 1.

$$ -1\leq Corr \leq 1$$
When we take Dist = 1- Corr
$$ 0\leq Dist = (1- Corr) \leq 2$$


##Evaluating cluster using Hierarchical Method


To evaluate the distance function, we need to extract the correlation and subtract to get the distance. 

An example for 10 genes

```{r clust1}
dist <- 1 - LPWC::corr.bestlag(simdata[11:20, ], timepoints = timepoints, max.lag = 2, penalty = "low", iter = 10)$corr
plot(hclust(dist))
```


These can also be assigned to cluster using cutree function. The example of cluster assignement with 3 cluster size

```{r clust 2}
dist <- 1 - LPWC::corr.bestlag(simdata[11:20, ], timepoints = timepoints, max.lag = 2, penalty = "low", iter = 10)$corr
cutree(hclust(dist), k = 3)
```



# Contributors

We have benefited in the development of LPWC from the help and
feedback of many individuals, including but not limited to: 

Gitter Lab Members,
James Dowell,
Ron Stewart


# Session info

```{r sessionInfo}
sessionInfo()
```


# FAQ

- How to speed up LPWC for low penalty with large gene set?
```{r, eval = FALSE}

comb <- function(x, ...) {
  lapply(seq_along(x),
         function(i) c(x[[i]], lapply(list(...), function(y) y[[i]])))
}



data <- simdata[1:10, ]
iter <- 10
allC <- findC(timepoints = timepoints, max.lag = 2)

##setting clusters
core <- parallel::detectCores() - 1
cl <- parallel::makeCluster(core)
doParallel::registerDoParallel(cl)


##results from multiple C
result <- foreach(i = 1:iter, .combine='comb', .multicombine=TRUE,
                .init=list(list(), list())) %dopar% {
            lags <- best.lag(data, max.lag = 3, timepoints = timepoints, C = allC[i])
            new.data <- prep.data(data = data, lags = lags, timepoints = timepoints)
            corr <- comp.corr(new.data$data, new.data$time, C = allC[i])
            return(list(corr, lags))
}
allcorr <- result[[1]]
alllags <- result[[2]]

##picking best C
val <- rep(NA, (length(iter) - 1))
for(i in 1:(iter - 1)){
  val[i] <- sum((as.vector(allcorr[[i + 2]]) - as.vector(allcorr[[i + 1]]))^2)
}

result <- list(corr = allcorr[[which.min(val) + 1]], lags = alllags[[which.min(val) + 1]], C = values[which.min(val) + 1])




```


